---
title: "Environmental Plastic Waste Analysis"
author: Shravani Subramanian
format: html
embed-resources: true
theme: cerulean
---

```{r} 
#| label: install-load-packages
#| message: false
library(tidyverse)
library(dplyr)
library(httr)
library(tidyjson)
library(jsonlite)
library(glue)
library(purrr)
library(countrycode)
library(forstringr)
library(scales)
library(hexbin)
library(broom)
library(gt)
```

## Context

The Environmental Plastic Waste Data comes from the **"Break Free from Plastic"** movement and covers information on plastic use across the world, organized by country, type of plastic, source of plastic, and year (2019 or 2020). The data comes from brand audits and cleanups conducted by volunteers. Some limitations of this dataset are that the countries will not have an equal sampling effort and the data depends on volunteer participation. Additionally, the amount of collected waste does not necessarily tell us how much plastic there is in general. Aside from those, the data is useful to help identify major corporate contributors to plastic pollution in our environment. It can also help inform environmental policy workers and provide corporate accountability.

## Data Cleaning

It seems as though initially the 2019 and 2020 datasets were imported separately from directories. Only the csv files were chosen from those respective directories. We then read the two CSV files with specified column types to prevent any confusion. The code then combines multiple files into a single file for each year (one for 2019 and one for 2020) through `map_dfr`. Next, the code cleans column names for both files using the `janitor::clean_names()` command, and makes the names consistent across years. A year column is also added to the datasets.

Individual Changes:

2020:

-   The `Grand_Total` variable is converted from character to numeric.

2019:

-   The country names are extracted from file names & stored in a `country` column.

-   Combines duplicate columns `pp_2` and `pp` as well as `ps` & `ps_2`.

-   Handles missing values using `is.na` and an `ifelse()` statement.

-   Drops `ps_2` & `pp_2`

Lastly, the 2019 and 2020 datasets are merged using `bind_rows()`and exported to a csv file.

## Research Questions

Using Current Data:

1)  Have certain plastic types (e.g. Polystyrene or High density polyethylene) become more or less common over time?


Using Supplemental Data:

1) How does annual mean temperature affect total plastic count? 

2) Do countries with higher GDPs have more plastic counting events, and thus a higher total plastic count?


# Checkpoint 3

## Temperature Data Set (Additional Country-Level Data)

```{r}
#| label: read-in-temp-data
# Read in Temperature data set
temp <- read.csv("combined_temperature.csv")

temp <- temp |>
  filter(Year %in% (c(2019, 2020)),) |>
  rename(AnnualMeanTemp = Annual.Mean)

# Show first 10 rows of temp data set
head(temp, 10)
```

The temperature data set includes data on the mean annual temperatures for all countries worldwide for the years 1901-2022, but we are particularly interested in the years 2019 and 2020. The 3-letter ISO code for each country is also included. We chose this data set because we wanted to explore how climate affects plastic degradation and consumption across different countries around the world.


## GDP Data Set (Additional Country-Level Data)

```{r}
#| label: gdp-data-api
#| eval: false
get_gdp <- function(country, key = my_key) {
  safe_parse <- safely(
    function(country, key) {

      response <- GET(
        url = "https://api.api-ninjas.com/v1/gdp",
        query = list(country = country),
        add_headers("X-Api-Key" = key)
      )

      data <- content(response, as = "text", encoding = "UTF-8")
      fromJSON(data)
    }
  )

  out <- safe_parse(country, key)

  # if there was an error, return empty data frame
  if (!is.null(out$error)) {
    return(data.frame())
  }

  # if result is empty, return empty data frame
  if (length(out$result) == 0) {
    return(data.frame())
  }

  # force result into data frame
  as.data.frame(out$result) |> filter(year %in% c(2019, 2020))
}

# Test Case
# can call function with 2-letter ISO code (get_gdp("CA")) or with full country name:
get_gdp("Canada")
```

```{r}
#| label: combine-lists-into-df
#| eval: false
gdp <- map_dfr(unique(temp$Code), get_gdp)
head(gdp, 10)
```

```{r}
#| label: write-csv-gdp
#| eval: false
write_csv(x = gdp, file = "stat431-gdp.csv")
```

The original GDP data set includes historical, current, and predicted data on GDP for countries around the world for the years for 1980-2029. We are particularly interested in the years 2019 and 2020. The 3-letter ISO code for each country is also included. There are 5 variables measuring GDP in different ways. For example, the variable `gdp_per_capita_nominal` measures GDP by dividing the country's total value of finished goods and services by its total population. `gdp_nominal` is the nominal GDP in billions of US dollars and `gdp_growth` expresses the GDP growth rate as a percentage. 

```{r}
#| label: read-csv-gdp
gdp <- read_csv("stat431-gdp.csv")
gdp
```


## "Meta" Dataset: Combining Temperature, GDP, and Plastic Waste Data Sets


```{r}
#| label: combine-temp-gdp-data
# Keep gdp_per_capita_nominal variable
gdp_per_capita <- gdp |>
  select(country, year, gdp_nominal, gdp_per_capita_nominal)

# Remove X5.yr.smooth variable from temperature data, join with gdp data set
combined_temp_gdp <- temp |>
  select(-X5.yr.smooth) |>
  left_join(gdp_per_capita, by = c("Code" = "country", "Year" = "year"))
```

```{r}
#| label: clean-plastics
plastics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2021/2021-01-26/plastics.csv')

# Sum up plastic and grand total plastic counts for each country, for each year
plastics_sum <- plastics |>
  group_by(country, year) |>
  summarize(
    across(empty:grand_total,\(x) sum(x, na.rm = TRUE)),
    num_events = first(num_events),
    volunteers = first(volunteers),
    .groups = "drop"
  )
head(plastics_sum, 10)
```

```{r}
#| label: country-code-plastics
# Create country code variable for plastics data
plastics_sum <- plastics_sum |>
  filter(!(country == "EMPTY")) |>
  mutate(Code = countrycode(sourcevar = country, origin = "country.name", destination = "iso3c")) |>
  rename(Year = year, Country = country)

head(plastics_sum, 10)
```

```{r}
# Combine plastics data with temp/GDP
meta_data <- inner_join(plastics_sum, combined_temp_gdp, by = c("Code", "Year")) |>
  select(-Country.y) |>
  rename(Country = Country.x)
```
```{r}
#| echo: false
#| results: false
meta_data$Country[meta_data$Country == "NIGERIA"] <- "Nigeria"
meta_data$Country[meta_data$Country == "ECUADOR"] <- "Ecuador"
meta_data$Country[meta_data$Country =="United Kingdom of Great Britain & Northern Ireland"] <- "United Kingdom"
meta_data$Country[meta_data$Country == "Cote D_ivoire"] <- "Cote Divoire"

head(meta_data, 10)
```

### How does annual mean temperature affect total plastic count? 

```{r}
#| label: data-summaries-viz-1
# Grand Total Plastic Count by Annual Mean Temperature
ggplot(data = meta_data, aes(x = AnnualMeanTemp, y = grand_total, color = factor(Year))) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) + 
  labs(x = "Annual Mean Temperature (Celsius)",
       y = "Grand Total Plastic Count",
       title = "Grand Total Plastic Count by Annual Mean Temperature",
       color = "Year") + 
  scale_color_manual(values = c("2019" = "steelblue", "2020" = "lightcoral")) + 
  theme_classic()
```

### Do countries with higher GDPs have more plastic counting events, and thus a higher total plastic count?

```{r}
#| label: data-summaries-viz-2
# Grand Total Plastic Count by Number of Plastic Counting Events
ggplot(data = meta_data, aes(x = num_events, y = grand_total, color = factor(Year))) +
  geom_point(aes(size = gdp_per_capita_nominal)) +
  labs(x = "Number of Plastic Counting Events",
       y = "Grand Total Plastic Count",
       title = "Grand Total Plastic Count by Number of Plastic Counting Events",
       color = "Year", 
       size = "Nominal GDP per Capital") + 
  scale_color_manual(values = c("2019" = "steelblue", "2020" = "lightcoral")) + 
  theme_classic()
```

```{r}
write_csv(x=meta_data, file = "stat431-meta.csv")
```


# Checkpoint 4

```{r}
#| label: read-in-meta-data
#| echo: true
#| output: false
plastic_temp_gdp <- read_csv("stat431-meta.csv")
```

```{r}
#| label: add-region-col
plastic_temp_gdp <- plastic_temp_gdp |>
  mutate(continent = countrycode(Country,
                                  origin = "country.name",
                                  destination = "continent"))
```

## RQ #1: How does change in GDP relate to change in the total amount of plastic collected?

```{r}
changed <- plastic_temp_gdp |>
  group_by(Country, continent, Year) |>
  summarise(grand_total = sum(grand_total, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = Year, values_from = grand_total) |>
  rename(
    total_2019 = `2019`,
    total_2020 = `2020`
  ) |>
  mutate(change = total_2020 - total_2019) |>
  drop_na(change, continent)
```

```{r}
#| fig-cap: Figure 1.1. Barplot for Change in Grand Total Plastic from 2019-2020, by Continent
#| fig-alt: Figure 1.1 Barplot showing changes in grand total plastic collection across continents between 2019 to 2020.

continent_year <- plastic_temp_gdp |>
  group_by(continent, Year) |>
  summarise(grand_total = sum(grand_total, na.rm = TRUE), .groups = "drop")

ggplot(continent_year, aes(x = continent, y = grand_total, fill = factor(Year))) +
  geom_col(position = "dodge") +
  labs(
    title = "Grand Total by Continent (2019 vs 2020)",
    x = "Continent",
    y = "Grand Total",
    fill = "Year"
  ) +
  scale_fill_manual(values = c("steelblue", "lightgreen")) +
  theme_classic() + 
  theme(plot.title = element_text(face = "bold"))
```



```{r}
#| fig-cap: Figure 1.2. Boxplot for Change in Grand Total Plastic from 2019-2020, by Continent
#| fig-alt: Figure 1.2. Boxplot showing change in grand total plastic collection across continents between 2019 and 2020. 
ggplot(changed, aes(x = continent, y = change)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.5) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Change in Grand Total by Continent",
    subtitle = "Country-level change from 2019 to 2020",
    x = "Continent",
    y = "Change (2020 - 2019)"
  ) +
  theme_classic() + 
  theme(plot.title = element_text(face = "bold"))
```

We noticed an outlier in Africa that was that was heavily compressing the scale of the boxplot. We identified the outlier to be Nigeria. 

```{r}
#| fig-cap: Figure 1.3. Boxplot for Change in Grand Total Plastic from 2019-2020, by Continent (excluding Africa outlier, Nigeria)
#| fig-alt: Figure 1.2. Boxplot showing change in grand total plastic collection across continents between 2019 and 2020, with Nigeria (outlier) removed.

ggplot(changed, aes(x = continent, y = change)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.5) +
  coord_cartesian(ylim = c(-30000, 30000))+
  scale_y_continuous(labels = comma) +
  labs(
    title = "Change in Grand Total by Continent",
    subtitle = "Country-level change from 2019 to 2020, excluding Nigeria",
    x = "Continent",
    y = "Change (2020 - 2019)"
  ) +
  theme_classic() + 
  theme(plot.title = element_text(face = "bold"))
```

```{r}
#| fig-cap: Figure 1.4. Barplot for Change in GDP from 2019-2020, by Continent
#| fig-alt: Figure 1.4. Barplot showing change in GDP across continent from 2019 to 2020.
continent_year <- plastic_temp_gdp |>
  group_by(continent, Year) |>
  summarise(gdp_nominal = sum(gdp_nominal, na.rm = TRUE), .groups = "drop")

ggplot(continent_year, aes(x = continent, y = gdp_nominal, fill = factor(Year))) +
  geom_col(position = "dodge") +
  labs(
    title = "Nominal GDP by Continent (2019 vs 2020)",
    x = "Continent",
    y = "Nominal GDP",
    fill = "Year"
  ) +
  scale_fill_manual(values = c("steelblue", "lightgreen")) +
  theme_classic() + 
  theme(plot.title = element_text(face = "bold"))
```


```{r}
# Calculating Percent Change in GDP from 2019 to 2020
gdp_changed <- plastic_temp_gdp %>%
  arrange(Country, Year) %>%
  group_by(Country) %>%
  mutate(gdp_pct_change = (gdp_nominal / lag(gdp_nominal) - 1) * 100)
```

```{r}
#| fig-cap: Figure 1.5. Boxplot for Percent Change in GDP from 2019-2020, by GDP
#| fig-alt: Figure 1.5. Boxplot showing percent change in GDP across continents from 2019 to 2020.
#| warning: false
ggplot(gdp_changed, aes(x = continent, y = gdp_pct_change)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +
  labs(y = "GDP % Change (2019-2020)", x = "Continent") + 
  theme_classic() + 
  theme(plot.title = element_text(face = "bold"))
```


## RQ #2: Is there a relationship between a country's mean annual temperature and the total amount of plastic collected?

```{r}
#| message: false 
#| fig-cap: Figure 2. Hexbin plot of Mean Annual Temperature and Grand Total Plastic Collected (2019-2020) by Continent
#| fig-alt: Figure 2. Hexbin plot showing mean annual temperature and grand total plastic collected from 2019 to 2020, by continent.

ggplot(data = plastic_temp_gdp) +
  geom_hex(mapping = aes(x = AnnualMeanTemp, y = grand_total)) +
  scale_fill_gradient(low = "lightblue", high = "royalblue4") +
  geom_smooth(aes(x = AnnualMeanTemp, y = grand_total), 
              method = "lm", se = FALSE, 
              color = "black", linewidth = 0.5, lty = 2) +
  labs(x = "Mean Annual Temperature (C°)", 
       y = "Grand Total Plastic Collected", 
       fill = "Number of Countries",
       title = "Number of Countries Observed", 
       subtitle = "For Mean Annual Temperature, Grand Total Plastic Collected Combination (2019 - 2020)") +
  theme_classic() + 
  theme(plot.title = element_text(face = "bold")) 
```

```{r}
# Fitting a linear model
plastic.fit <- lm(grand_total~AnnualMeanTemp, data = plastic_temp_gdp)
summary(plastic.fit)
```
Heatmap of Plastic Composition by Continent
```{r}
plastic_long <- plastic_temp_gdp %>%
  pivot_longer(
    cols = c(hdpe, ldpe, o, pet, pp, ps, pvc),
    names_to = "plastic_type",
    values_to = "amount"
  )
heatmap_data <- plastic_long %>%
  group_by(continent, plastic_type) %>%
  summarise(total_amount = sum(amount, na.rm = TRUE), .groups = "drop")

heatmap_prop <- heatmap_data %>%
  group_by(continent) %>%
  mutate(prop = total_amount / sum(total_amount))

ggplot(heatmap_prop, aes(x = plastic_type, y = continent, fill = prop)) +
  geom_tile(color = "white") +
  scale_fill_gradientn(colors = c("#03045E","#00B4D8", "#CAF0F8"),
    labels = scales::percent
  ) +
  labs(title = "Plastic Composition by Continent",
       fill = "Proportion",
       x= "Plastic Type", 
       y= "Continent") +
  theme_minimal()
```

Stacked Barplot for Plastic Type Composition by Continent
```{r}
continent_plastic <- plastic_long %>%
  group_by(continent, plastic_type) %>%
  summarise(total_amount = sum(amount, na.rm = TRUE), .groups = "drop")

ggplot(continent_plastic, aes(x = continent, y = total_amount, fill = plastic_type)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c(
    "hdpe" = "#809bce",
    "ldpe" = "#95b8d1",
    "o"    = "#b8e0d2",
    "pet"  = "#d6eadf",
    "pp"   = "#bde0fe",
    "ps"   = "#a2d2ff",
    "pvc"  = "#eac4df" 
    ), 
    labels = c(
    "hdpe" = "HDPE",
    "ldpe" = "LDPE",
    "o" = "OTHER",
    "pet" = "PET",
    "pp" = "PP",
    "ps" = "PS",
    "pvc" = "PVC")
    ) +
  labs(
    title = "Plastic Type Composition by Continent",
    x = "Continent",
    y = "Proportion",
    fill = "Plastic Type"
  ) +
  theme_minimal()
```
Scatterplot of Log Grand total and Temperature
```{r}
ggplot(plastic_temp_gdp, aes(x = AnnualMeanTemp, y = grand_total, color = continent)) +
  geom_point(alpha = 0.7, size = 3) +
  scale_y_log10(labels = scales::comma) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    title = "Temperature vs Plastic Collection (Log Scale)",
    x = "Mean Annual Temperature (°C)",
    y = "Total Plastic Collected (Log Scale)",
    color = "Continent"
  ) +
  theme_minimal()
```

```{r}
rq1_long <- plastic_temp_gdp %>%
  filter(Year %in% c(2019, 2020)) %>%
  pivot_longer(
    cols = c(hdpe, ldpe, o, pet, pp, ps, pvc),
    names_to = "plastic_type",
    values_to = "amount"
  )
rq1_summary <- rq1_long %>%
  group_by(continent, plastic_type) %>%
  summarise(
    total_amount = sum(amount, na.rm = TRUE),
    .groups = "drop"
  )

rq1_prop_by_type <- rq1_summary %>%
  group_by(plastic_type) %>%
  mutate(prop = total_amount / sum(total_amount)) %>%
  ungroup()

rq1_wide_by_type <- rq1_prop_by_type %>%
  select(continent, plastic_type, prop) %>%
  pivot_wider(
    names_from = plastic_type,
    values_from = prop
  )

```
```{r}
rq1_wide_by_type %>%
  gt() |>
  tab_header(
    title = "Proportion of Each Plastic Type by Continent",
    subtitle = "Totals combined across 2019 and 2020"
  ) |>
  data_color(
    columns = c(hdpe, ldpe, o, pet, pp, ps, pvc),
    fn = col_numeric(
      palette = c("#CAF0F8", "#0077B6"),  # light → dark blue
      domain = c(0, 1)
    )) |>
  tab_style(
    style = cell_text(weight = "bold", color = "black"),
    locations = list(
      cells_title(groups = "title"),          
      cells_column_labels(everything())
    )
  ) |>
  fmt_percent(
    columns = c(hdpe, ldpe, o, pet, pp, ps, pvc),
    decimals = 1
  ) |>
  cols_label(
    continent = "Continent",
    hdpe = "HDPE",
    ldpe = "LDPE",
    o = "Other",
    pet = "PET",
    pp = "PP",
    ps = "PS",
    pvc = "PVC"
  )
```


